\documentclass[a4paper,11pt]{article}

\usepackage[latin1]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{bbm} %math chars
\usepackage{amsmath}
\usepackage{indentfirst}
\usepackage{fullpage} %minimizes the default margins
\usepackage{url}
\usepackage{graphicx}
\usepackage[center,footnotesize]{caption} %options des legendes des graphes
\usepackage[section]{placeins} %place les figures d'une section avant le debut de la suivante
\usepackage{subfig} %a) b) c)

\title{Series 5}
\date{}
\author{Genomics and bioinformatics - Week 6 - September 23, 2012}

\begin{document}
\maketitle

This series is about phylogenetic trees and a complement to HMMs.

\section{Phylogenetic trees}

Circadian rhythms (often referred to as the ``body clock``) are
ubiquitous in most life forms. The ``period`` family of genes is an 
essential component of this clock.

The table below provides BLAST scores of pairwise
alignments for six period family proteins from three different
species, 

\begin{enumerate}
\item {\bf dm\_per} from {\it Drosophila melanogaster}, i.e. fruit fly,
\item {\bf dr\_per2} and {\bf per3} from {\it Danio rerio}, i.e. zebra fish,
\item {\bf mm\_per1}, {\bf mm\_per2} and {\bf mm\_per3} from {\it Mus
  musculus}, i.e. mouse.
\end{enumerate}

\begin{center}
	\begin{tabular} {| l || l | l | l | l | l | l |}
	\hline
 	& dm\_per & dr\_per2 & dr\_per3 & mm\_per1 & mm\_per2 & mm\_per3 \\ \hline\hline
	dm\_per & -  & 116 & 125 & 129 & 123 & 121 \\	\hline
	dr\_per2 &	114 & - & 661 & 887 & 947& 497 \\	\hline
	dr\_per3 & 123 & 644 & - & 667 & 666 & 699 \\	\hline
	mm\_per1 & 126 & 745 & 572 & - & 640 & 80 \\	\hline
	mm\_per2 & 126 & 984 & 680 & 832 & - & 455 \\	\hline
	mm\_per3 & 124 & 499 & 709 & 537 & 488 & - \\
	\hline
	\end{tabular}
\end{center}
\vspace{0.05 cm}

\begin{enumerate}
\item Based on the table above, draw the gene tree for the six period
  proteins (dm\_per, dr\_per2, dr\_per3,  mm\_per1, mm\_per2,
  mm\_per3). 
\item Using this tree, give examples of:
  \begin{itemize} 
  \item an orthologous pair, 
  \item a paralogous pair in the same species, and 
  \item a paralogous pair in different species.
  \end{itemize}
\end{enumerate}

\section{Another HMM}

\subsection{The Viterbi algorithm}

... fill a 2x4 grid by hand

\newpage

\subsection{Using R}

In this exercise we show you how to use an existing package to model a problem with an HMM and solve it. There exist such packages for most programming languages, but here we arbitrarily chose R.

\subsubsection{Get started}

Start R, then download, install and import the HMM package:

\begin{verbatim}
install.packages("HMM")
library(HMM)
\end{verbatim}

Create an array of hidden states names \texttt{states}, an array of emission names \texttt{emissions}, a matrix \texttt{transProb} with the transition probabilities, a matrix \texttt{emissionProb} 
with the emission probabilities, and a vector \texttt{initProb} with the probabilities of the initial state:

\begin{verbatim}
states <- c("N","I")
emissions <- c("A","T","C","G")
initProb <- c(0.5, 0.5)
transProb <- matrix(c(0.9, 0.45, 0.1, 0.55), 2)
emissionProb <- matrix(c(0.25, 0.05, 0.25, 0.05, 0.25, 0.45, 0.25, 0.45), 2)
\end{verbatim}

Create the model with these parameters:

\begin{verbatim}
hmm <- initHMM(states, emissions, initProb, transProb, emissionProb)
hmm  # prints a summary of the model
\end{verbatim}

Using the model created above, let us generate a random observation of length 100:

\begin{verbatim}
simulation <- simHMM(hmm,100)
simulation  # prints the chosen hidden states and emissions sequences
\end{verbatim}

\subsubsection{Visualize}

Copy and paste the following code to visualize the situation. The plot on top shows the observed sequence,
that at the bottom shown the hidden states:

\begin{verbatim}
obsToVal <- function(obs){
	if((obs == "A") or (obs=="T")){return(2)}
	else {return(4)}
}
stateToVal <- function(st){
	if(st == "I"){return (1)}
	if(st == "N"){return (3)}
	else {return ("NA")}
}
par(mfrow=c(2,1))
plot( sapply(simulation$observation, obsToVal),pch=19,ylim=c(0,8),col=sapply(simulation$observation, obsToVal),cex=.5)
legend("topright",c("A/T","G/C"),pch=19,col=c("red","blue"))
plot( sapply(simulation$states, stateToVal),type="l",ylim=c(0,5))
\end{verbatim}

\subsubsection{Forward, Backward algorithms}

Here are the commands to run the forward and backward algorithms. What is the probability of the observed sequence, given the model? :

\begin{verbatim}
logForwardProb <- forward(hmm, simulation$observation)
exp(logForwardProb)
logBackwardProbabilities <- backward(hmm, simulation$observations)
exp(logBackwardProb)
\end{verbatim}

\subsubsection{Viterbi}

For the simulated sequence, we knew the hidden states. Now say we have an observed sequences, but
do not know the hidden states. Try first to guess where are the ``I'' region(s) in these two sequences:

\begin{verbatim}
obs1 <- c("A","T","G","A","A","G","T","A","C","T","T","A","A","G","T","G","A",
          "C","T","A","A","T","T","A","G","T","T","T","A","T","C","T","A","A")
obs2 <- c("A","T","A","G","A","T","T","C","T","G","C","C","G","A","A","A","T",
          "T","T","C","A","T","T","A","G","T","T","T","A","T","C","T","A","A")
\end{verbatim}

Now we are going to find the most probable hidden sequence that explains these observations thanks to the Viterbi algorithm. Is it close to what you expected?

\begin{verbatim}
viterbi1 <- viterbi(hmm, obs1)
viterbi1
viterbi2 <- viterbi(hmm, obs2)
viterbi2
\end{verbatim}

\subsubsection{Baum-Welsch}

Finally, let us admit that our transition and emission probabilities are just a first guess, but we would like to update the model parameters by training it with the Baum-Welsch algorithm:

\begin{verbatim}
bw = baumWelch(hmm, obs2, 10)  # 10 iterations
print(bw$hmm)
\end{verbatim}

\end{document}











